---
title: "Homework - Econometrics I"
subtitle: São Paulo School of Economics - FGV/EESP
author: 
  - "Student: Rafael Felipe Bressan" 
  - "Professor: Bruno Ferman"
date: "`r Sys.Date()`"
output: 
  bookdown::html_book:
    highlight: espresso
    toc: false
    number_sections: false
bibliography: "econometrics.bib"
biblio-style: "apalike"
link-citations: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Questions

@Angrist1998 consider two different instrumental variables to estimate the causal effects of family size on parents’ labor supply.

**1. Explain why it would be important to have a valid instrument in this case. You expect that OLS would be biased in which direction?**

The paper's authors aim to explain the **causal** effects of childbearing on labor-supply of mothers mainly, but also of husbands (supposedly the father). On the economic side, fertility would have a negative impact on mother's labor-supply, since, as the theory goes, mother's would cut on work hours to stay at home and provide childcare. On the other hand, demographers have been testing theories about how mothers wages and labor force attachment impact on fertility.

Since fertility cannot be at the same time exogenous and determined by measures of labor-supply, it must be the case that the two are **jointly determined**, causing endogeneity in any regression attempt of labor-supply on fertility and no causal inference can be made of this kind of regression.

This type of **endogeneity** is called by @Hansen2020 as simultaneous equation bias to make it clear that both variables are being jointly determined by a system of equations that must be solved simultaneously.

In order to have non-biased estimate of childbearing effect on labor-supply one of the strategies that the researcher can make use of is **Instrumental Variables - IV**. But even if the instrument is valid, this still does not ensure us the estimated coefficient has a causal interpretation, this comes only when the instruments considered, _same sex_ and _twins-2_ are in fact randomly assigned to each mother, since no one can choose the sex of the baby or whether to have a single or multiple births. Only when the instruments chosen are randomly assigned that the authors are able to claim their regression is assessing causal effects of childbearing on family labor-supply.

Although the paper explicitly says "jointly determined", one can approach the problem as a regression with omitted explanatory variable and as such, it presents a *omitted variable bias* if OLS is to be employed. We can think of "desire to have a couple" as the omitted variable that has some explanatory power over labor supply. Let's denote this unobservable variable as $c$ and we have that the bias from not taking it at account in our regression model is:

$$\hat{\beta_{ols}}-\beta=\delta \frac{cov(x, c)}{var(x)}$$

where $\beta$ is the true parameter relating labor-supply to fertility, $x$, and $\delta$ is the parameter relating the desire of having a couple as children to labor-supply.

It's natural to suppose that $\delta<0$, since parents, and especially mothers with a great desire to have children of opposing gender are more concerned about their family, home environment and childbearing than of making a professional career. Meanwhile, the covariance of $x$ and $c$ is positive, that is, couples trying to have kids of different sex tend to have more children. They need to play the odds, plain and simple.

Therefore, one can suppose that OLS bias in this setup is negative and it actually **exacerbates the negative** relation of fertility to labor-supply.

**2. Put yourself in the shoes of the authors of the paper. How would you justify the validity of each of these two instrumental variables?**

In order to justify the validity of any instrumental variable the researcher needs to attest two conditions: relevance of the instrument and the exclusion condition.

Relevance of the instrument says that our potentially endogenous regressor variable, $x$, has to be correlated to our instrument of choice, $z$. That is, the expected value of the product of these variables should not be equal to zero, $\mathbb{E}[z_i x_i]\neq 0$, for all individuals, $i$, in our population. Since the regressor chosen is a dummy variable that indicates that the mother has more than two children, the instruments _same_sex_ and _twins-2_ are correlated to the regressor. The authors, following an extensive literature on the matter, argument that parents who have the first two kids of the same sex are more prone to have a third one, while the parents of twins at the second pregnancy get the third kid as a surprise.

The exclusion condition is held when the instrument affects the dependent variable, $y$, _only_ through its effect on the regressor, that is, there is no other channel of transmission, direct or indirect, in which the instrument influences the dependent variable. This means that the expected value of the product of the instrument and the structural equation error term is equal to zero, $\mathbb{E}[z_ie_i]=0$ when the structural equation is given by: $y_i=\mathbf{x_i'\beta}+e_i$. For the _same_sex_ instrument it is easy to see that, since having the first two kids is random, it should not have any direct influence on the labor-supply provision, both for mother and father. As for the _twins-2_ goes, things are not that clear it is not dependent on other mother's demographic characteristics, but anyway, having twins is largely a random variable and no one expects it to have any impact on the choice of labor-supply.

**3. Now put yourself in the shoes of a referee evaluating the paper. Discuss different potential threats to the validity of these instruments.**

First start with _twins-2_ instrument. There is some literature that relates having twins to maternal characteristics like age, height, weight, race and the use of fertility treatments, @Farbmacher2018 cites a few of them. Also, having dizygotic twins is largely influenced by genetic traits, @Painter2010, and so the conditional expectation of having twins for a woman who is first degree related to a dizygotic twin, or is a twin herself, is not the same as the general population. Women that undergo a fertility treatment are making an endogenous decision that we know raises the probability of having twins, and these treatments are evermore common nowadays. 

The _twins-2_ indicator is constructed by the authors, it is not a mother's declaration that their kids are in fact twins. This poses some issues with the instrument created for example, the case where the husband is the father of one the children and the kids happen to be about the same age (i.e. the same quarter). In this case the children would be deemed as twins, an assumption far from reality. Moreover, the fact that a single mother happens to find a suitable husband, father of a kid of the same ages as hers, might influence on the decision to get married and the marriage can have an impact on labor-supply decision. 

All these facts contribute to invalidate the _twins-2_ as an instrument to fertility and its randomness, spoiling the causal inference that otherwise could be done.

TODO: ARGUMENT AGAINST SAME SEX.  

Male children have poorer health at birth and thus mothers with two boys are more prone to stay at home or have medical appointments more often then families of two girls. Therefore the instrument might have a direct impact on labor-supply and thus, invalidating the exclusion hypothesis, this instrument would be invalid.

**4. Discuss how the importance of the threats you discussed in the previous item might vary if you consider the same instruments for different countries or different years.**

The threat of _twins-2_ be an invalid instrument seems to be more relevant as the time passes, since modern families structure are going towards households with children from different parents marriages living under the same roof ever more common. Fertility treatments are now common place and many women postpone childbearing in favor of developing their professional careers. When they decide to have a child, the age factor might influence in the necessity of fertility treatment and thus raising the probability of having twins. Especially on the Western cultures, this seems to be affecting almost all countries. For other religion/culture countries we lack information about this topic and cannot make any assertion.

Public policies that provide incentives to give birth or otherwise, like China.

TODO: ARGUMENT ABOUT SAME SEX.

**5. Select a country and a census year from [IPUMS](https://international.ipums.org/international/index.shtml). Estimate the effects of family size on parents’ labor supply using this sample. Each student should choose a different sample. Once you select your sample, sign up in the google doc to guarantee no one else is using the same sample [(link)](https://docs.google.com/spreadsheets/d/1tI9acQk0LdPMhXMwQoVlOex1fnxsqmf8-anfytZoZgc/edit?usp=sharing). Submit a well-polished table with the results. Analyse how your results differ from @Angrist1998 . Submit your codes in an appendix.**

This regression model is 


\begin{equation}
y_i=\alpha_0'w_i+\alpha_1 s_{1i}+\alpha_2 s_{2i}+\beta x_i+\varepsilon_i
(\#eq:structural)
\end{equation}

\begin{equation}
x_i=\pi_0'w_i + \pi_1 s_{1i} + \pi_2 s_{2i} + \gamma same\_sex_i+\eta_i
(\#eq:first-stage)
\end{equation}

where $w_i$ are the demographic control variables, mother's age and age at first birth, plus race and Hispanic indicators. The control variables for sex are $s_{1i}$ and $s_{2i}$ for first and second child respectively, which equal one if the child's gender is male.

Present Tables 6 and 7 of the paper.

```{r libraries, message=FALSE}
library(AER)
library(lmtest)
library(sandwich)
library(tidyverse)
library(broom)
library(stargazer)
```

```{r random-gen, cache=TRUE}
n_sample <- 1000
set.seed(123)
s1 <- sample(c(0,1), size = n_sample, replace = TRUE)
s2 <- sample(c(0,1), size = n_sample, replace = TRUE)
two_boys <- s1*s2
two_girls <- (1 - s1)*(1 - s2)
same_sex <- two_boys + two_girls
# Tilt the probability of having more than two children if same sex
probs <- ifelse(same_sex, 0.6, 0.4)
more_two_child <- as.numeric(rbernoulli(n_sample, p = probs))
# Test routine
# mean(more_two_child[which(same_sex == 1)])
# mean(more_two_child[which(same_sex == 0)])
# Now define labor_income to be negatively dependent on more_two_child
set.seed(123)
labor_income <- pmax(10000 - 1850*more_two_child + rnorm(n_sample, sd = 1000), 0)
df <- data.frame(labor_income = labor_income,
                 s1 = s1,
                 s2 = s2,
                 more_two_child = more_two_child,
                 same_sex = same_sex,
                 two_boys = two_boys,
                 two_girls = two_girls)
```

```{r first-stage}
first_reg <- lm(more_two_child ~ same_sex, data = df)
first_robse <- sqrt(diag(vcovHC(first_reg, type = "HC1")))
```

```{r iv-reg}
ols_reg <- lm(labor_income ~ s1 + s2 + more_two_child, data = df)
iv_reg <- ivreg(labor_income ~ s1 + s2 + more_two_child | s1 + s2+ same_sex, data = df)

ols_coef <- coeftest(ols_reg, vcov. = vcovHC, type = "HC1")
iv_coef <- coeftest(iv_reg, vcov. = vcovHC, type = "HC1")
```

```{r table6, results='asis'}
notes6 <- "Robust standard error in parenthesis."
stargazer(first_reg, type = "html", header = FALSE,
          label = "tab:table6",
          title = "First stage estimates of more than 2 children equation.",
          dep.var.caption = "",
          dep.var.labels = "labor income",
          covariate.labels = c("same sex", "constant"),
          report = "vcs",
          se = list(first_robse),
          omit.stat = c("rsq", "adj.rsq", "ser"),
          omit.table.layout = NULL,
          notes = notes6,
          notes.append = FALSE)
```

**6. Using the data from the previous item, how could you provide evidence in favor and/or against the validity of the instruments?**

## References {-}