---
title: "Problem Set 3 - Econometrics I"
subtitle: São Paulo School of Economics - FGV/EESP
author: 
  - "Student: Rafael Felipe Bressan" 
  - "Professor: Bruno Ferman"
date: "`r Sys.Date()`"
output: 
  bookdown::html_book:
    highlight: tango
    toc: false
    number_sections: false
bibliography: "econometrics.bib"
biblio-style: "apalike"
link-citations: true
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(haven)
library(sandwich)
library(AER)
library(car)
library(broom)
library(kableExtra)
library(stargazer)
library(tidyverse)
```

## Question 5

**During the 1880s, a cartel known as the Joint Executive Committee (JEC) controlled the rail transport of grain from the Midwest to eastern cities in the United States. The cartel preceded the Sherman Antritrust Act of 1890, and it legally operated to increase the price of grain transportation above what would have been the competitive price. From time to time, cheating by members of the cartel brought about a temporary collapse of the collusive price-setting agreement. In this exercise, you will use variations in supply associated with the cartel’s collapses to estimate the elasticity of demand for rail transport of grain. We will use a data file JEC.dta that contains weekly observations on the rail shipping price and other factors from 1880 to 1886. So, our main goal is to estimate the elasticity of demand for rail shipping of grain. In particular, we wish to regress $\ln(Q_i)$ where $Q_i$ is the total tonnage of grain shipped in week $i$, on $\ln(P_i)$, where $P_i$ is the price of shipping a ton of grain by rail.**

**(a) Estimate the price elasticity of demand by using OLS to regress the log of the quantity of grain shipped on the log of the price and the full set of monthly binary indicators. What is the estimated value of the demand elasticity and its standard error?**

Loading Dataset

```{r load, cache=TRUE}
JEC <- read_dta("JEC.dta")
data_red <- data.frame(ln_price = log(JEC$price), 
                       ln_quantity = log(JEC$quantity), 
                       JEC[,5:16])
```

The OLS regression model is:

\begin{equation}
\ln(Q_i)= \beta_0 + \beta_1 \ln(P_i)+\sum_{j=1}^{12}\gamma_j seasj_i + \varepsilon_i\\
(\#eq:ols-reg)
\end{equation}

```{r ols-reg, results = "asis"}
reg_ols <-  lm(ln_quantity ~ . , data = data_red)
se_ols <- sqrt(diag(vcovHC(reg_ols, type = "HC1")))

stargazer(reg_ols, type = "html", header = FALSE,
          label = "tab:table1",
          title = "OLS estimation.",
          dep.var.caption = "",
          dep.var.labels = "Log-quantity",
          report = "vcs",
          se = list(se_ols),
          omit.stat = c("rsq", "adj.rsq", "ser"),
          omit.table.layout = "n",
          omit = "seas*"
          )
```


Estimated value of price elasticity of demand is `r coef(reg_ols)["ln_price"]` , while its standard error is `r se_ols["ln_price"]`.

**(b) Explain why the interaction of supply and demand could make the OLS estimator of the elasticity obtained in (a) biased.**

Supply and demand compose what we know as **market equilibrium**. Quantity and price are jontly determined in a system of equations and we can only observe market equilibrium, never demand or supply isolated. This jointly determination of variables causes OLS to be biased, something that @Hansen2020 calls _simultaneous equations bias_.

**(c) Consider using the variable cartel as instrumental variable for $\ln(P)$. Use economic reasoning to argue whether cartel plausibly satisfies the two conditions for a valid instrument.**

From economic theory we known that cartel formation can interfere on prices. Companies secretly agrees to charge a higher than competitive markets prices in order to boost their profits, thus the relevance of the instrument is clear.

What is not without some argument is the exclusion condition. Cartels can, and often do, agree on quantities to be produced by members and therefore this instrument is not valid in this case. Specifically on the present situation, we can assume that supply for rail transportation of grains clears the demand, that is, all grains are transported and none is left in stock for long periods of time. Hence, the collusion is not affecting quantities supplied but rather only the prices that grain owners must accept to pay for transport.

Including the cartel data on the data set.

```{r cartel}
data_red$cartel <- JEC$cartel
```

Estimating the IV regression.

```{r iv-reg, results='asis'}
reg_iv <- ivreg(ln_quantity ~ ln_price + . - cartel | . - ln_price + cartel,
                data = data_red)
se_iv <- sqrt(diag(vcovHC(reg_iv, type = "HC1")))

stargazer(reg_iv, reg_ols, type = "html", header = FALSE,
          label = "tab:table2",
          title = "IV and OLS estimation compared.",
          # dep.var.caption = "",
          dep.var.labels = "Log-quantity",
          #covariate.labels = c("same sex", "constant"),
          report = "vcs",
          se = list(se_iv, se_ols),
          omit.stat = c("rsq", "adj.rsq", "ser"),
          omit.table.layout = "n",
          omit = "seas*"
          #notes = notes6,
          #notes.append = FALSE
          )
```

Utilizing `cartel` as an instrument variable for price, the estimated value of price elasticity of demand is `r coef(reg_iv)["ln_price"]`, while its standard error is `r se_iv["ln_price"]`.

If the instrument is valid and our regression is consistent, then the estimated **OLS bias** is `r coef(reg_ols)["ln_price"] - coef(reg_iv)["ln_price"]`.

**(d) Test, at 5% significance level, the hypothesis that the demand for rail shipping of grain is inelastic.**

Demand for rail shipping of grain is inelastic if its price elasticity is greater than -1, thus our hypothesis test is one-sided.

```{r test1, results = "asis"}
# Auxiliary function to perform t-test
# H0 beta_hat >= beta
ttest <- function(reg, coefnum, val, unilateral = FALSE){
  co <- coef(summary(reg))
  tstat <- (co[coefnum,1] - val)/co[coefnum,2]
  if (unilateral)
    pval <- pt(tstat, reg$df.residual, lower.tail = FALSE)
  else
    pval <- 2*pt(abs(tstat), reg$df.residual, lower.tail = FALSE)
  
  return(data.frame(t_value = tstat, p_value = pval))
}

# ln_price is the second coeficient
test1 <- ttest(reg_iv, 2, -1, unilateral = TRUE)
```

Our t-stat is `r test1$t_value` with a p-value of `r test1$p_value`.

**(e) Test, at 5% significance level, the hypothesis that the demand elasticity is equal to −1. What is the p-value of this test?**

We can use a asymptotic chi-squared Wald test for a linear restriction test, or the usual two-sided t-test.

```{r test2, results = "asis"}
test2_chi <- linearHypothesis(reg_iv, "ln_price = -1", 
                              vcov. = vcovHC, type = "HC1")
test2_t <- ttest(reg_iv, 2, -1)

df <- tribble(~Test,  ~Statistic, ~`p-value`,
              "t-test", test2_t$t_value, test2_t$p_value,
              "Wald", test2_chi$Chisq[2], test2_chi$`Pr(>Chisq)`[2])

kable(df, format = "html",
      digits = 4,
      booktabs = TRUE) %>% 
  kable_styling(full_width = FALSE,
                position = "left")

```

**(f) One of the reasons for price fluctuations was that the Great Lakes periodically froze, making shipping grain by boat impossible and temporarily increasing the demand for rail. $Ice_i$ is a binary variable that is equal to 1 if the Great Lakes are not navigable because of ice. Consider using the variable Ice as instrumental variable for $\ln(P)$. Use economic reasoning to argue whether Ice plausibly satisfies the two conditions for a valid instrument to estimate demand elasticity for rail shipping of grain.**

$Ice_i$ is actually a demand shifter and should be included in the demand equation. This exogenous variable would help to estimate the parameters from _supply equation_, but would be of no use as an instrument of price when estimating the demand equation.

Eventhough $Ice_i$ is clearly correlated to price, this is a result of its direct impact on quantity demanded and thus the exclusion hypothesis is not satisfied.

## Question 6

**In this exercise, we will study the impact of eminent domain on house prices in the US^1 . Eminent domain (or taking laws) is the legal mechanism through which a government can compulsorily purchase private property. This action is surely controversial, and proprietors can take legal action in order to try to revert such a decision. Federal court rulings declaring that a government takeover was unlawful (pro-plaintiff decisions) uphold property rights and set a precedent that makes future exercise of eminent domain more difficult. Our analysis will thus centre on the impact of pro-plaintiff rulings in federal circuits 2 on housing prices.**

**(a) Import temp `CSHomePrice.mat.csv`. This dataset stores information on house prices (Case-Shiller index) and pro-plaintiff rulings per circuit and period. Note that, except for the log CS index, all data is on a yearly frequency. Aggregate the dataset by year, taking the within-circuit-year average of the log price index.**

Loading the dataset and aggreting it by circut/year.

```{r load-cs, message=FALSE, cache=TRUE}
CS_home_price <- read_csv("temp_CSHomePrice.mat.csv")

# aggregate by year
cs_year <- CS_home_price %>% 
  group_by(circuit, year) %>% 
  summarise_all(mean)
```

**(b) Run a regression of the log-price index on the total number of pro-plaintiff appellate taking decisions (`numpro_casecat_12`). What do you find? Does this coefficient have a causal interpretation? Why?**

```{r reg1, results="asis"}
reg1 <- lm(logpriceindex ~ numpro_casecat_12, data = cs_year)
vcov_reg1 <- vcovHC(reg1, type = "HC1")
se_reg1 <- sqrt(diag(vcov_reg1))

notes1 <- "Robust standard error in parenthesis."
stargazer(reg1, type = "html", header = FALSE,
          label = "tab:table1",
          title = "OLS estimate of Case-Shiller home price index.",
          dep.var.caption = "",
          dep.var.labels = "log(price index)",
          covariate.labels = c("pro-plantiff", "Intercept"),
          report = "vcs",
          se = list(se_reg1),
          omit.stat = c("rsq", "ser"),
          omit.table.layout = NULL,
          notes = notes1,
          notes.append = FALSE)
```

blalblabla

**To deal with potential endogeneity, Chen and Yeh (2015) explore the fact that, in federal appellate courts, judges are randomly assigned to three-judge panels to decide appellate cases. Each Circuit Court case receives three _randomly_ assigned judges out of a pool of judges, numbering roughly 8 to 40 depending on the size of the Circuit. Therefore, conditional on the pool characteristics, the realized characteristics of the randomly assigned three-judge panels should satisfy the exclusion restriction, as they can only be related to property prices through the judges’ decisions.**



**(c) Match dataset `probs.txt` to your data. This dataset stores the distribution of characteristics in each circuit pool from 1945 to 2007.**

```{r probs, message=FALSE, cache=TRUE}
probs <- read_table2("probs.txt")

# Merge two datasets by left join cs_year and probs.
# Select the appropriate variables for next questions
cs_probs <- cs_year %>% 
  ungroup() %>% 
  left_join(probs, by = c("circuit" = "Scircuit", "year" = "Syear")) %>% 
  select(logpriceindex, numpro_casecat_12, missing_cy_12, numcasecat_12, 
         numpanels1x_dem, starts_with("prob"), -starts_with("prob_x"))
```

**(d) We propose to run the following model:**

$$\ln(Case\_Shiller)_{ct} = \beta_0 + \beta_1 numpro\_casecat\_12_{ct} + \gamma'W_{ct} +  \epsilon_{ct}$$

**where $W_{ct}$ includes the full set of controls for judicial pool characteristics; a dummy for whether there were no cases in that circuit-year (`missing_cy_12`), and the number of takings appellate decisions (`numcasecat_12`). As an instrument, we choose the number of panels with at least one Democrat (`numpanels1x_dem`).**

**Estimate the above model using the proposed instrument. Does the instrument satisfy the relevance assumption? Comment your estimates.**

```{r iv, results = "asis"}
ols_reg <- lm(logpriceindex ~ numpro_casecat_12+missing_cy_12+numcasecat_12+.-numpanels1x_dem,
              data = cs_probs)
iv_reg <- ivreg(logpriceindex ~ numpro_casecat_12+missing_cy_12+numcasecat_12+.-numpanels1x_dem |
                  numpanels1x_dem+missing_cy_12+numcasecat_12+.-numpro_casecat_12,
                data = cs_probs)
vcov_iv <- vcovHC(iv_reg, type = "HC1")
se_iv <- sqrt(diag(vcov_iv))

notes2 <- "Robust standard error in parenthesis."
stargazer(coeftest(iv_reg), ols_reg, type = "html", header = FALSE,
          label = "tab:table2",
          title = "IV estimate of Case-Shiller home price index.",
          dep.var.caption = "",
          dep.var.labels = "log(price index)",
          covariate.labels = c("pro-plantiff", "missing", "numcasecat",
                               "Intercept"),
          report = "vcs",
          se = list(se_iv, NULL),
          omit = c("prob_*"),
          omit.stat = c("rsq", "ser"),
          omit.table.layout = NULL,
          notes = notes2,
          notes.append = FALSE)
```

**(e) Do you think the coefficient found in (d) is a credible estimate of the causal effect we set ourselves to analyse? Why? If the answer is negative, could you use the data available and run an alternative specification that would produce more credible estimates? What did you find?**